What should I notice? Evaluating memorability of events for abductive inference.

### 2021-09-17 paper.tex
commit

As a user has just turned on the TV in her all-equipped living-room, the
lights dim and the window blinds lower. Intrigued by this behavior, she
quickly infers that both light dimming and the blinds closing occurred as a
consequence of the TV being turned on. How did she come to this conclusion?
By performing \emph{abductive inference} \cite{magnani_abduction_2011}, which
is a key element of human's ability to understand the world: from the
observed consequences, infer the possible causes.

@@

To tackle this challenge, we rely on the following observation: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of qualifiers, the most memorable ones are
likely to require less words to be described. Think, for instance, of how
``last year's hottest day'' and ``the 182nd day from 7 years ago'': the
former seems simpler to describe. By evaluating the length of each
description, taking into account both the complexity of concept words (a date
of occurrence, a temperature ranking), and the arguments (the hottest, the
182ndThis
kind of inference, named \emph{abductive inference} \cite
{magnani_abduction_2011}, is a key element of humans's ability to understand the world., 7) used in both descriptions, we can assign each event a \emph{description complexity}
score. Memorable events, as they stand out, would then differ from their
neighbors and require simpler and less numerous qualifiers to be
unambiguously described. As such, they will stand out as simpler.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\linewidth]{figures/filters}
  \caption{An example of retrieving an event through successive predicative
   filters. From the base memory (yellow), successive filters select the
   events satisfying the associated predicate (grey arrow). For example,
   filter $f_1$ selects events from last year, i.e. which satisfy the
   predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
   applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
   year's hottest day. The complexity of this event can then be upper bounded
   by the complexity of the three filters, since they give a unambiguous way
   to describe the event in the memory.}
   % their predicates (gray arrows). The
   % operations yields successive subsets of the memory (yellow), until a final
   % singleton memory is reach, meaning that the successive filters retrieve a
   % unique event from the memory. The predicates used for this retrieval path
   % and their arguments, $\pi_{1}, \pi_{2}, \pi_{3}, k_{1}, k_{2}, k_{3}$, can
   % be used as a retrieval program. Its length gives an upper bound to the
   % description complexity of the event $e_{h}$: ``last year's hottest
   % day''.}
  \label{fig:filters}
\end{figure}

For machines to use description complexity to distinguish memorable events from others, we need a formal canvas and computation methods which coincide with human intuition . To do so, we rely on principles from Algorithmic Information Theory, which also appear to be consistent with the human perception of complexity\cite{li_introduction_2008,dessalles2011coincidences,delahaye_numerical_2012}. Our method is as follows: we
consider events as being unordered elements from a base \emph{memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to an event and
eventual arguments. For instance, the predicate $\mathtt{date}
(\cdot, 1\_year)$ assigns \emph{true} to events that occurred last year. Selecting
all events, from the memory, that satisfy a given predicate corresponds to
a \emph{filter} operation, and yields another memory, a subset of the former.
This filtering operation can then be repeated, each iteration, selecting
fewer events, until a singleton memory is reached, meaning that the sequence
of predicates can \emph{retrieve} unambiguously the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this evaluation for the event
$e_h$: ``last year's hottest day''.

@@

The rest of this paper is organized as follows: we first briefly introduce
some notions of Algorithmic Information Theory and its applications in
section~\ref{sec:theory}, then we explicit our methods of computing the
unexpectedness of past events in cyber-physical systems~\ref
{sec:computing}. We illustrate this approach by providing an implementation
relying on a few predicates on events from a smart home simulation~
(section \ref{sec:example}). In section~\ref{sec:related} we briefly review
other related works using complexity theory or trying to explain smart homes
and how our work can be linked with them before exploring potential
extensions of our work in section \ref{sec:future}

@@

It is important to note that the notion of Kolmogorov complexity is not related
to computation complexity, as there is no requirement on the execution time of
the programs, only their length in bits matters for the computation of
complexity. In fact, it can be shown that Kolmogorov complexity is not
computable\cite{li_introduction_2008}. The sketch of the proof is the
following: since the definition of complexity from eq. \ref
{eq:kolm_def} requires to find the shortest of all programs on a given Turing
machine outputting $s$, it requires to run all programs up to a given length
and compare their result to $s$. However, being able to do so means that we
would be able to tell whether these programs terminate. Since the termination
of programs is famously undecidable, by extension, the computation of
Kolmogorov complexity is impossible. One can however easily approximate it with
upper bounds, by exhibiting a program outputting $s$.

@@

This bridge between Algorithmic Information Theory and human perception of
complexity is extended to define the notion of simplicity and unexpectedness,
which are considered of uttermost importance in cognitive science\cite
{chater_simplicity_2003}.
\cite{dessalles2011coincidences} proposes a formal definition of the
 unexpectedness $U(e)$ of an event, as the difference between an a-priori
 expected causal complexity $C_{w}(e)$ and the actual observed complexity $C
 (e)$.
\begin{equation}
  \label{eq:unexpected}
  U(e) = |C_{w}(e) - C(e)|
\end{equation}
This result comes from the understanding that, while Kolmogorov complexity is
computed with Turing machine, it can be used as an elegant proxy for modeling
information processing in the human brain, and thus helps design a notion of
simplicity or complexity of events.

Using this measure, it is possible to model phenomenons such as coincidences:
imagine that you happen to run into someone in a park. If this person is not
noticeable, the event will be quite trivial. On the other hand, this person is
your best friend, as the description complexity $C(e)$ drops, the causal
complexity $C_{W}(e)$ remains unchanged, and the event becomes noticeable.

Using these insights from AIT, we propose to measure the complexity of past
events $e$ from a memory $\mathcal{M}$ by considering the quantity
$K(e|\mathcal{M})$, that is the length of the shortest program outputting $e$,
given the entire memory of past events $\mathcal{M}$ as an input. To limit the
search and obtain a computable measure, we limit ourselves to only a subset of
possible programs, which we will now describe.

\section{Computing the unexpectedness of events}
\label{sec:computing}
\subsection{Retrieving an event}
We model the memory as an unordered set $M$ of events $e$. Individual events can
be seen as data points, augmented with a label indicating their nature
(temperature events, failure events, addition/removal events), and a timestamp:
\begin{equation}
  \label{eq:event}
  e = (l, t,\mathcal{D})
\end{equation}
where $l$ is the label, $t$ a timestamp and $\mathcal{D}$ a multi-dimensional
data point representing various characteristics of the events (e.g.~its duration,
the maximum temperature reached, the sensor name, its position).

We define \emph{predicates} as boolean function operating on these events, and
taking an additional argument, encoded as a finite binary string:
\begin{equation}
  \label{eq:predicate}
  \pi : \begin{cases}
    \mathcal{M}\times \{0,1\}^{*} &\mapsto \{0,1\} \\
    (e, k) &\mapsto \pi_{k}(e)
    \end{cases}
\end{equation}
An example of predicate is to use $\pi = \mathtt{year}$ and $k=1$, thus
constructing the predicate $\mathtt{year()}e, Y\mathtt{)}$, which tells whether
the event $e$ occurred $Y$ years ago.

From these predicates, we can construct filter operations on the memory. A
filter operates by selecting the events satisfying a given predicate
$\pi_{k}$ and constructing a resulting memory from these events.
\begin{equation}
  \label{eq:filter}
 f_{\pi, k}(\mathcal{M}) = \left\{e \in \mathcal{M} | \pi_{k}(e) \right\}
\end{equation}
For instance, using the same $\pi = \mathtt{year}$ and $k=1$ as above, we would
build the filter $f_{\pi, k} = \mathtt{last\_{}year}$, which selects all events
that occurred last year.

As the output of a filter applied to the memory $\mathcal{M}$ is another memory
object $\mathcal{M'} = f_{\pi, k}(\mathcal{M})$, we can compose filter
functions. A sequence of such filters, applied in succession, is called a
\emph{retrieval path}
\begin{equation}
\label{eq:ret_def}
p = (f_{\pi_{1}, k_{1}}, \dots, f_{\pi_{n}, k_{n}})
\end{equation}
and by definition
$p(\mathcal{M}) = f_{\pi_{n}, k_{n}}(\dots(f_{\pi_{1}, k_{1}}(\mathcal{M})))$.
In case the result of the operation $p(\mathcal{M})$ contains a single element
$e$, we say that the path $p$ \emph{retrieves} the element $e$ from $\mathcal{M}$, and write
$p(\mathcal{M}) = e$.

@@

% From the previous definitions, we then estimate the complexity of events as the
% ``simplest'' way of retrieving them from the memory, by applying successive
% filters. A retrieval path composed of various filters
% $p = f_{1} = f_{\pi_{1}, k_{1}}, f_{2} = f_{\pi_{2}, k_{2}}, \dots, f_{n} = f_{\pi_{n}, k_{n}}$
% is said to \emph{retrieve} an event $e$ from a memory $\mathcal{M}$ if and only
% if
% $p(\mathcal{M}) = f_{1} \circ f_{2} \circ \dots \circ f_{n} (\mathcal{M}) = \{e\}$.
% For an event $e$, if such a path exist, its complexity can be bounded by
% $K(p) = \sum_{f_{\pi, k} \in p} l(\pi) + l(k)$, the complexity of the retrieval
% path being the sum of the bit lengths required to describe the predicates $\pi$
% and their arguments $k$.

As presented in sec. \ref{sec:theory}, we are interested in computing the
Kolmogorov complexity of an event $e$, conditionned on the memory $\mathcal{M}$.
As this quantity is not computable, we use the previous definitions of retrieval
paths and approximate the quantity $K(e|\mathcal{M})$ by the lenght of the
shortest retrieval path $p$ retrieving $e$ from $\mathcal{M}$.

Given that all predicate concepts can be described by a prefix program of length
$l(\pi)$, and their arguments in length $l(k)$, each retrieval path $p$ can be
described by the sequence of prefix description of all the predicate concepts and
arguments used to construct its filters:

\begin{equation}
K(p) = l((f_{\pi_{1}, k_{1}}, \dots, f_{\pi{n}, k_{n}})) = \sum_{i=0}^{n} l(\pi_{i}) + l(k_{i})
\end{equation}

Using this definition of the lenght of a retrieval path, we can compute the
estimation for the complexity $K(e|\mathcal{M})$, which, for clarity, we will
denote simply by $K(e)$ for the rest of this paper.

\begin{equation}
  \label{eq:desc_k}
  K(e) \approx \min_{p | p(\mathcal{M}) = e} K(p) = \min_{p | p(\mathcal{M})=e} \sum_{f_{\pi, k} \in p} l(\pi) + l(k)
\end{equation}

@@

The definition used for description complexity in eq. \ref{eq:desc_k} allows for
a direct implementation, which is shown in alg. \ref{alg:complex_iter}. This
algorithm operates iteratively: starting with the base memory $\mathcal{M}$
(line 1), we apply all possible predicate concepts $\pi$ from a given set
$\mathcal{P}$ and programs $k$ (lines 6-7), up to a given length $\mathtt
{max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
{explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explore retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).

\subsection{Computing an unexpectedness measure}
Equation \ref{eq:unexpected} gives a universal definition of unexpectedness, as
the difference between the complexity of the phenomenon generating an event and
the complexity required to describe this event. In our particular application
case, we may directly use the description complexity $K(e)$ as defined in
\ref{eq:desc_k}.

The term $C_{w}(e)$ from eq. \ref{eq:unexpected} is harder to formally
fit into our canvas. As $C_{w}$ is defined as the shortest program a
``W-machine'' replicating the world's behavior, based on the person's
knowledge\cite{dessalles2011coincidences}, it is convenient to understand this
term as the expectation one would have regarding the complexity of the event
$e$, \emph{before} actually evaluating it.

A possible proxy for this quantity is to compute the average complexity
of \emph{similar} events. Still, difficulty remains in the definition of what
can be considered \emph{similar} events. Given that we have already used the
notion of predicates to describe the events and compute their description
complexity, we can use them as simple metrics for similarity: for each predicate
$\pi$ satisfied by $e$, we can define a set of similar events $N_{\pi}(e)$ as
all the events also satisfying $\pi$.

\begin{equation}
\label{eq:similar}
N_{\pi}(e) = \{e'\in \mathcal{M}, \pi(e') \wedge e' \neq e\}
\end{equation}

Now, when considering, for all possible predicates $\pi$, the corresponding
neighborhoods $N_{\pi}(e)$, with the convention that $N_{\pi}(e) = \emptyset$
if $\pi(e) = \bot$, we can compute an average expected complexity for $e$:

\begin{equation}
\label{eq:expected}
K_{exp}(e) = \frac{
                    \sum_{\pi} \sum_{e' \in N_{\pi}(e)} K(e')
                  }{
                    \sum_{\pi} |N_{\pi}(e)|
                  }
\end{equation}

The definition used for the expected complexity of the event $e$ in eq.
\ref{eq:expected} holds the intuition of favoring more similar events: if $e'$
is \emph{very} similar to $e$, it will appear in many neighborhoods, since it
satisfies mostly the same predicates as $e$. Therefore, its complexity term will
be present in more terms in eq. \ref{eq:expected}, and will weight more towards
the final result.

\subsection{From memorability to abduction}
% On peut étendre la définition utilisée pour la surprise en faisant intervenir
% les prédicats'' `sympathiques` par rapport à ce que l'on questionne.
The problem of the abductive inference is different from the basic computation
of the memorability score. Here, \emph{knowing} that we want to find a cause $c$
to an observed effect $e$, we try to find the most remarkable event in past
memory in that regards. While our ``memorability'' score aims to identify the
most remarkable events, it does not take into account the added knowledge about
the effect $e$ that is available for abductive inference.

This added knowledge would however influence the unexpectedness of a possible
cause $c$, from eq. \ref{eq:unexpected}: even though the causal complexity
$C_{W}(c)$ is likely to remain unchanged, as it relies on the model of the world in
the person's mind, the description complexity $C(c)$ is replaced with the
conditional complexity $C(c|e)$, as the knowledge of $e$ is included in the
problem's data.

In our canvas of predicative filtering, this corresponds to adding for free, in
terms of program length, information about $e$ to predicates. This translates
into considering relative order between events: when investigating a cause for
an anomaly in the living-rooms, other anomalies occurring in the same
living-room will be simpler, since the location ``living-room'' is already known
from knowing the consequence. In terms of predicates, we therefore append a
description of $e$ to all programs $k$ of predicates:
$\pi_{k}(c) = \pi_{k::e}(c)$, where $::$ is the append operation.

This leads to another score $Abd(c,e)$ (abduction score of effect $c$ for cause
$e$), which is defined similarly to the description complexity in eq.
\ref{eq:desc_k} and using relative filters.

\begin{equation}
  \label{eq:abd_k}
  Abd(c) = \min_{p | p(\mathcal{M}|e) = c} K(p) = \min_{p | p(\mathcal{M}|e)=c} \sum_{f_{\pi, k} \in p} l(\pi) + l(k)
\end{equation}
This abduction score translates the additional information provided to the
system when answering a user's request. As such, it encapsulates the idea
presented as the motivation of this paper: when confronted to a surprising
situation, and with no other source of information, it can be used as a metrics
to find relevant possible surprising causes to the observed consequence.

@@

\begin{figure}[ht]
  \includegraphics[width=\linewidth]{figures/ts_example}
  \caption{Time series data from the simulation: outdoor temperature (blue) and
    controller temperature of a room (blue). On different occasion, the
    controlled temperature deviates from its setting (21°C). We will evaluate if
  our system finds these deviations as memorable, and if it can propose causal
  hypotheses.}
  \label{fig:ts_example}
\end{figure}

@@

For the implementation of our method, we first needed to identify and
characterize events from the time series data generated by the iCasa simulation.
Since this is not the focus point of our present work (see sec.
\ref{sec:related}), we simply apply threshold and pre-computed conditions based
detection to create a base of events.

This base of events consists the basis of the initial memory $\mathcal{M}$ used
for computations.
We then implemented a small number of predicate concepts:
\begin{itemize}
        \item \texttt{label(e, k)}: whether the event $e$ has the label ranked
        $k^{th}$ in the memory.
        \item \texttt{rank(e, k)}: whether the event $e$ has the rank $k_{1}$
        along axis $k_{2}$, where $k$ is a unique encoding for $(k_{1}, k_{2})$.
        \item \texttt{day(e, k)}: whether the event $e$ occurred $k$ days ago.
        \item \texttt{month(e, k)}: whether the event $e$ occurred $k$ months
        ago.
        \item \texttt{location(e, k)}: whether the event $e$ occurred in zone
        $k$.
\end{itemize}

@@

The results of the description complexity evaluation, for the described setup,
are shown in fig. \ref{fig:computed_cplx}. The entire computation, over 4
iterations (meaning that retrieval paths contained at most 4 filters), took
around 30 seconds on a modern commercial laptop with an i7-8700u CPU.

@@

% TODO
\textbf{TODO}

>>>>

As a user has just turned on the TV in her all-equipped living-room, the
lights dim and the window blinds lower. Intrigued by this behavior, she
quickly infers that both light dimming and the blinds closing occurred as a
consequence of the TV being turned on. How did she come to this conclusion?
By performing \emph{abductive inference} \cite{magnani_abduction_2011}, which
is a key element of humans' ability to understand the world: from the
observed consequences, infer the possible causes.

@@

To tackle this challenge, we rely on the following observation: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of qualifiers, the most memorable ones are
likely to require less words to be described. Think, for instance, of
``last year's hottest day'' and ``the 182nd day from 7 years ago'': the
former seems simpler to describe. To quantify the notion, we can evaluate the complexity of each
description, taking into account both the complexity of concept words (a date
of occurrence, a temperature ranking), and the arguments (the hottest, the
182nd, 7). The resulting value defines the  \emph{description complexity}
of the events. As our intuition is that memorable events require simpler and less numerous qualifiers to be
unambiguously described than other unremarkable ones, they will stand out as simpler in terms of description complexity.

@@

\begin{figure}[ht]
  \centering
\includegraphics[width=\linewidth]{figures/filters}
\caption{Retrieving an event through successive predicative
filters. From the base memory (yellow), successive filters select events satisfying the associated predicate (grey arrows). For example,
   filter $f_1$ selects events from last year, i.e. which satisfy the
   predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
   applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
   year's hottest day. The complexity of this event can then be upper bounded
   by the complexity of the three filters, since they give a unambiguous way
to describe the event in the memory.}
  \label{fig:filters}
\end{figure}

For machines to implement and compute description complexity, we need a formal canvas and computation methods which coincide with human intuition . To do so, we rely on Algorithmic Information Theory, which also appears to be consistent with the human perception of complexity\cite{li_introduction_2008,dessalles2011coincidences,delahaye_numerical_2012}. Our method is as follows: we
consider events as being unordered elements from a base \emph{memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to an event and
eventual arguments. For instance, the predicate $\mathtt{date}
(\cdot, 1\_year)$ assigns \emph{true} to events that occurred last year. Selecting
all events, from the memory, that satisfy a given predicate corresponds to
a \emph{filter} operation, and yields another memory, a subset of the former.
This filtering operation can then be repeated, each iteration, selecting
fewer events, until a singleton memory is reached, meaning that the sequence
of predicates can \emph{retrieve} unambiguously the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this evaluation for the event
$e_h$: ``last year's hottest day''.

@@

The rest of this paper is organized as follows: we first briefly introduce
some notions of Algorithmic Information Theory and its applications in
section~\ref{sec:theory}, then we explicit our methods to compute the
memorability of past events in cyber-physical systems~\ref
{sec:computing}. We illustrate this approach by providing an implementation
relying on a few predicates on events from a smart home simulation~
(section \ref{sec:example}). In section~\ref{sec:related} we briefly review
other related works using complexity theory or trying to explain smart homes
and how our work can be linked with them before exploring potential
extensions of our work in section \ref{sec:future}

@@

It is important to note that the notion of Kolmogorov complexity is not related
to computation complexity, as there is no requirement on the execution time of
the programs, only their length in bits matters for the computation of
complexity. In fact, it can be shown that Kolmogorov complexity is not
computable\cite{li_introduction_2008}. The sketch of the proof is the
following: since the definition of complexity from Equation. \ref
{eq:kolmo_def} requires to find the shortest of all programs on a given Turing
machine outputting $s$, it requires to run all programs up to a given length
and compare their result to $s$. However, being able to do so means that we
would be able to tell whether these programs terminate. Since the termination
of programs is famously undecidable, by extension, the computation of
Kolmogorov complexity is impossible. One can however easily approximate it with
upper bounds, by exhibiting a program outputting $s$.

@@

This bridge between Algorithmic Information Theory and human perception of
complexity is extended to define the notion of simplicity and unexpectedness,
which are considered of uttermost importance in cognitive science\cite
{chater_simplicity_2003}.
\cite{dessalles2011coincidences} proposes a formal definition of the
 unexpectedness $U(e)$ of an event, as the difference between an a-priori
 expected causal complexity $C_{w}(e)$ and the actual observed complexity $C
 (e)$.
\begin{equation}
\label{eq:unexpected} U(e) = C_{w}(e) - C(e)
\end{equation}
This result comes from the understanding that, while Kolmogorov complexity is
computed with Turing machine, it can be used as an elegant proxy for modeling
information processing in the human brain, and thus helps design a notion of
simplicity or complexity of events.

The definition of Equation \ref{eq:unexpected} allows to model phenomenons
such as coincidences: imagine that you happen to run into someone in a park.
If this person has no particular link to you, the event will be quite
trivial: the complexity of describing this person will be equivalent to
distinguishing it from the global population, which will be roughly
equivalent as the complexity of describing that this person happens to be in
the same park as you. On the other hand, if you run into your best friend in
a park, as the complexity of describing your best friend is significantly
lower, the description complexity $C(e)$ drops while the causal complexity
remains $C_W (e)$ unchanged. Therefore the event becomes unexpected.

Using these insights from AIT, we define the memorability $M(e)$ of an event as the absolute difference between the description complexity $K_d(e)$ of an event and its expected description complexity $K_{exp}(e)$:
\begin{equation}
  \label{eq:memorability}
  M(e) = |K_{exp}(e) - K_d(e)|
\end{equation}
Contrary to the definition of unexpectedness from Equation \ref{eq:unexpected}, we use an absolute value: this is done to acknowledge events more complex than expected as memorable\footnote{In the original paper \cite{dessalles2011coincidences}, unusually complex events could be handled by considering complexity itself as a way to describe the event: see ``the Pisa Tower effect''\cite{dessalles_pisa_nodate}}. In the next section, we formally define the description complexity $K_d$ and the expected complexity $K_{exp}$ of events and detail how we can compute approximations of these quantities.

\section{Computing the memorability of events}
\label{sec:computing}
\subsection{Retrieving an event}

Computing the memorability of events begins with the introduction of formal definitions of events, how they are stored, how we can describe them and how these descriptions can be used to retrieve them.

We define \emph{events} as data points augmented with a \emph{label} indicating their nature (temperature event, failure event, addition/removal of a device) and a timestamp of occurrence. Formally:
\begin{equation}
  \label{eq:event}
  e = (l, t,\mathcal{D})
\end{equation}
where $l$ is the label, $t$ the timestamp and $\mathcal{D}$ a multi-dimensional data point representing the various characteristics of $e$: its duration, the maximum temperature reached, the sensor name, its position, etc. Labels can also be considered as classes of events, of which each event is a particular instance.

To model how humans are able to describe events by using qualifiers, we use \emph{predicates}: function operating on events, eventually taking additional parameters, returning a boolean value: $\pi(e, a_1, a_2, \dots, a_n) \mapsto \{O,1\}$ is a predicate of arity $n$ operating on event $e$. In the rest of this paper, we will prefer the equivalent notation $\pi(e, k) \mapsto \{0,1\}$, where $k$ is a binary string encoding the sequence of arguments $a_1, \dots, a_n$. Using this notation, the predicate $\pi$ becomes a boolean function operating on $\mathbf{E} \times \{0,1\}^*$:
\begin{equation}
  \label{eq:predicate}
  \pi : \begin{cases}
    \mathcal{M}\times \{0,1\}^{*} &\mapsto \{0,1\} \\
    (e, k) &\mapsto \pi_{k}(e)
    \end{cases}
\end{equation}
An example of predicate is to take $\pi = \mathtt{year}$ and $k$ a string encoding the number $1$, thus
constructing the predicate $\mathtt{year(}e, 1\mathtt{)}$, which tells whether
the event $e$ occurred $1$ years ago.

As events occur, they are stored in a \emph{memory} $M_0$. As they are not directly comparable, the memory $M_0$ can be considered as having the structure of an unordered set. We denote by $\mathcal{M}$ the subsets of $M_0$. By extension, elements of $\mathcal{M}$, i.e. subsets of $M_0$, are also called \emph{memories}.

By applying a given predicate $\pi_k$ to all events contained in a memory $M \subseteq M_0$, and selecting only events satisfying $\pi_k$ yields another memory $M_1 \subseteq M \subseteq M_0$. We call this operation a \emph{filter}:
\begin{equation}
  \label{eq:filter}
f_{\pi, k}: \begin{cases}
  \mathcal{M} & \mapsto \mathcal{M}             \\
  M           & \mapsto \{e \in M | \pi_k(e) \}
\end{cases}
\end{equation}
For instance, using the same $\pi = \mathtt{year}$ and $k=1$ as above, we can
build the filter $f_{\pi, k} = \mathtt{last\_{}year}$, which selects all events
that occurred last year.

As the output of a filter applied to a memory $M$ is another memory
object $M' \subseteq M$, we can compose filter
functions. A sequence of such filters is called a \emph{retrieval path}
\begin{equation}
\label{eq:ret_def}
p = (f_{\pi_{1}, k_{1}}, \dots, f_{\pi_{n}, k_{n}})
\end{equation}
and by definition
$p(M) = f_{\pi_{n}, k_{n}}(\dots(f_{\pi_{1}, k_{1}}(M)))$.
In case the result of the operation $p(M)$ contains a single element
$e$, we say that the path $p$ \emph{retrieves} the element $e$ from $M$, and write
$p(M) = e$. In the example shown in Figure \ref{fig:filters}, the three filters $f_1, f_2, f_3$ form a retrieval path retrieving the event ``last year's hottest day'' from the base memory $M_0$.

@@

As presented in sec. \ref{sec:theory}, we are interested in computing an approximation of the description complexity of an event $e$. From the above definitions, if there is a path $p$ retrieving $e$ from the base memory $M_0$, i.e. $p(M_0) = e$, this path provides a possible unambiguous description for $e$. We therefore define the description complexity of $e$ as the minimum complexity of a path $p$ retrieving $e$ from the base memory $M_0$.

\begin{equation}
  \label{eq:k_desc}
  K_d(e) = \min_{p \in P_\infty} \{L(p) | p(M_0) = e\}
\end{equation}

where the bit-length $L(p)$ of a retrieval path is defined as the number of bits of a string encoding the path. If we limit ourselves to prefix-free strings encoding predicates and arguments, the total bit length is given by:
\begin{align}
  \label{eq:bit_lenght_p}
  L(p) & = L((f_{\pi_1,k_1}, \dots, f_{\pi_n, k_n}))     \\
       & = L(\pi_1) + L(k_1) + \dots + L(\pi_n) + L(k_n)
\end{align}

By considering only a finite number of possible predicates $\pi$ and arguments $k$, and a maximum path length, we can construct a finite set $P$ of possible retrieval paths. By limiting the search over this set, we can upper bound the description complexity, and use this upper bound as an approximation:

\begin{equation}
\label{eq:approx_k_desc}
K_d(e) \leq \min_{p \in P \land p(M_0) = e} L(p) = \min_{p \in P \land p(M_0)=e} \sum_{f_{\pi, k} \in p} L(\pi) + L(k)
\end{equation}

@@

The approximation of description complexity from Equation \ref{eq:approx_k_desc} allows for
a direct implementation, which is shown in Algorithm \ref{alg:complex_iter}. This
algorithm operates iteratively: starting with the base memory $M_0$
(line 1), we apply all possible predicate concepts $\pi$ from a given finite set
$\Pi$ and programs $k$ (lines 6-7), up to a given length $\mathtt
{max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
{explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explore retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).

\subsection{Computing Memorability}
As stated in Equation \ref{eq:memorability}, we define memorability $M(e)$ as the absolute difference between the description complexity of an event and its expected value. As we've just defined $K_d(e)$ and provided an approximation in Equation \ref{eq:approx_k_desc}, we now focus on defining the description expected complexity of an event, $K_{exp}(e)$.

As intended in the Equation \ref{eq:memorability}, this term evaluates the complexity the user, or the system, would expect the event $e$ to have without actually describing it, based on their previous knowledge. In our canvas, this prior knowledge consists of the base memory $M_0$. The expected complexity of the event $e$ can be computed with a simple first-order approximation, i.e. estimating the average complexity of ``similar events'' over the base memory $M_0$.

Still, difficulty remains in the definition of what
can be considered \emph{similar} events. Given that we deal with non comparable events, the only possibility to universally define the notion of similarity is to once again refer to \emph{predicates}. Thus, for a given event $e$, and a given predicate $\pi_k$, we define a $\pi_k$ neighborhood of $e$ as the set $N_{\pi, k}(e)$ of all other events satisfying $\pi_k$.

\begin{equation}
\label{eq:similar}
N_{\pi, k}(e) = \{e'\in M_0, \pi_k(e') \wedge e' \neq e\}
\end{equation}

Now, when considering, for all possible predicates $\pi_k$, the corresponding
neighborhoods $N_{\pi, k}(e)$, with the convention that $N_{\pi, k}(e) = \emptyset$
if $e$ does not satisfy $\pi_k$, we can compute an average expected complexity for $e$:

\begin{equation}
\label{eq:expected}
K_{exp}(e) = \frac{
\sum_{\pi, k} \sum_{e' \in N_{\pi, k}(e)} K_d(e')
                  }{
\sum_{\pi, k} |N_{\pi, k}(e)|
                  }
\end{equation}

This definition is consistent with the intuitive ideas that more similar events should weight more in the computation. Indeed, if $e'$
is \emph{very} similar to $e$, it will appear in many neighborhoods, since it
satisfies mostly the same predicates as $e$. Therefore, it will
be present in more terms in Equation \ref{eq:expected}, and will weight more in
the final result.

\subsection{From memorability to abduction}
% On peut étendre la définition utilisée pour la surprise en faisant intervenir
% les prédicats'' `sympathiques` par rapport à ce que l'on questionne.
The problem of the abductive inference is different from the basic computation
of the memorability score. Here, \emph{knowing} that we want to find a cause $c$
to an observed effect $e$, we try to find the most remarkable event in past
memory in that regard. While our ``memorability'' score aims to identify the
most remarkable events, it does not take into account the added knowledge about
the effect $e$ that is available for abductive inference.

This added knowledge can be integrated into the description complexity definition by using conditioned predicates. That is considering as given, and therefore ``free'' in terms of complexity, information contained in the cause effect $e$. For instance, when looking for a cause for
an anomaly in the living-room, other anomalies occurring in the same
living-room will be simpler, since the location ``living-room'' is already known
from knowing the consequence.

Formally, we now consider only predicates where knowledge of the effect event $e$  is appended to all programs $k$:
$\pi_{k::e}(c)$, where $::$ is the append operation. The set of paths obtained with such predicates is $P^\infty_e$. This append operation is free in terms of bit-length in the computation of complexity, since the effect event $e$ is an input of the problem. Therefore, we have $L'(\pi_{k::e}) = L(\pi_k) = L(\pi) + L(k)$. Which finally gives the definition for the conditional description complexity:

\begin{align}
  \label{eq:abd_k}
  K_d(c | e) & = \min_{p \in P^\infty_e} \{L'(p), \quad p(M_0)=c \}                                                \\
             & = \min_{p \in P^\infty_e} \left\{\sum_{f_{\pi, k::e} \in p} L(\pi) + L(k), \quad p(M_0) = c\right\}
\end{align}

This new conditional description complexity translates the additional information provided to the
system when answering a user's request. It can then be averaged over similar events to compute the expected conditional description complexity, $K_{exp}(c|e)$. From this, we come to the definition of the conditional memorability, which measures how memorable an event $c$ is, considering the knowledge of another event $e$:

\begin{equation}
\label{eq:cond_mem}
M(c|e) = |K_{exp}(c|e) - K_d(c|e)|
\end{equation}

Conditional memorability encapsulates the idea
presented as the motivation of this paper: when confronted to a surprising
situation, and with no other source of information, we can identify events that appear more memorable than others, with regards to the target effect event. As such, our conditional memorability score provides a ranking that can be used for abductive inference, proposing as best candidates the most memorable ones.

@@

\begin{figure}[ht]
  \includegraphics[width=\linewidth]{figures/ts_example}
  \caption{Time series data from the simulation: outdoor temperature (blue) and
controller temperature of a room (orange). On different occasion, the
    controlled temperature deviates from its setting (21°C). We will evaluate if
  our system finds these deviations as memorable, and if it can propose causal
  hypotheses.}
  \label{fig:ts_example}
\end{figure}

@@

For the implementation of our method, we first needed to identify and
characterize events from the time series data generated by the iCasa simulation.
Since this is not the focus point of our present work (see sec.
\ref{sec:related}), we simply apply threshold and pre-computed conditions based
detection to create a set of events.

This set of events constitutes the basis of the initial memory $M_0$ used
for computations.
We then implemented a small number of predicate concepts:
\begin{itemize}
\item $\mathtt{label(e, k)}$: whether the event $e$ has the label ranked
        $k^{th}$ in the memory.
\item $\mathtt{rank(e, a_1, a_2)}$: whether the event $e$ has the rank $a_1$
along axis $a_2$.
\item $\mathtt{day(e, k)}$: whether the event $e$ occurred $k$ days ago.
\item $\mathtt{month(e, k)}$: whether the event $e$ occurred $k$ months
        ago.
\item $\mathtt{location(e, k)}$: whether the event $e$ occurred in zone
        $k$.
\end{itemize}

@@

The results of the description complexity evaluation, for the described setup,
are shown in fig. \ref{fig:computed_cplx}. The entire computation, over 4
iterations (meaning that retrieval paths contained at most 4 filters), took
around 30 seconds on a commercial laptop with an i7-8700u CPU.

@@

\textbf{TODO: the experiment still needs to be done: some additional data is required here (creating the scenario and finding causes.)}

\subsubsection{Discussion}
\label{ssub:discussion}
